\documentclass[sigconf]{acmart}

% ── Packages ──────────────────────────────────────────────────────────────────
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{url}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows, positioning, fit}

% ── Metadata ──────────────────────────────────────────────────────────────────
\title{FakeShield: An End-to-End MLOps Pipeline for Real-Time Fake News Detection with Automated Drift Monitoring and HPC-Triggered Retraining}

\author{Ahmed Zegdane}
\affiliation{%
  \institution{[Your University]}
  \department{[Department]}
  \city{[City]}
  \country{[Country]}
}
\email{[your.email@university.edu]}

\begin{abstract}
The proliferation of misinformation on social media demands automated, continuously adaptive detection systems.
Existing fake news detection research predominantly focuses on static benchmark evaluations, neglecting the
deployment challenges of production systems — namely concept drift, domain mismatch, and the operational
overhead of retraining. We present \textit{FakeShield}, an end-to-end Machine Learning Operations (MLOps)
pipeline that spans the full lifecycle of a fake news detection system: data ingestion, multi-model training,
real-time social media monitoring, statistical drift detection, and automated GPU retraining on a national
high-performance computing (HPC) cluster. We conduct a systematic comparison across five model families
— TF-IDF with Logistic Regression and SVM, LSTM, TextCNN, and BERTweet — trained on the FakeNewsNet
benchmark, and further fine-tune BERTweet on the PHEME rumour dataset for tweet-domain adaptation.
Our best model achieves an AUC-ROC of 0.882 on FakeNewsNet. The pipeline continuously monitors
live political discourse from Twitter/X, computes Population Stability Index (PSI) and Kolmogorov-Smirnov
statistics against the training distribution, and automatically triggers SLURM-scheduled retraining on
TRUBA GPU nodes when drift is detected. FakeShield demonstrates that production-grade MLOps practices
can be applied to misinformation detection with modest infrastructure cost.
\end{abstract}

\keywords{fake news detection, MLOps, drift monitoring, BERTweet, continuous training, PHEME}

% ──────────────────────────────────────────────────────────────────────────────
\begin{document}
\maketitle

% ── 1. INTRODUCTION ───────────────────────────────────────────────────────────
\section{Introduction}

Misinformation on social media platforms poses serious risks to public discourse, democratic processes,
and public health~\cite{vosoughi2018spread}. While the NLP community has produced strong fake news
detection models~\cite{zhou2020survey}, a critical gap persists between benchmark performance and
real-world deployment: models are trained once on a static dataset, then deployed into an ever-changing
information landscape without monitoring or adaptation.

This deployment gap manifests in two ways. First, \textit{concept drift} — the statistical properties of
live social media data shift continuously as new events, political figures, and narratives emerge. Second,
\textit{domain mismatch} — models trained on curated news headlines (e.g., GossipCop, PolitiFact) are
often applied to short social media posts with very different linguistic characteristics. The result is
silent model degradation that practitioners have no automated mechanism to detect or correct.

We address this gap with \textit{FakeShield}, a complete MLOps pipeline built around four core principles:

\begin{enumerate}
  \item \textbf{Reproducibility}: every experiment is tracked with DVC and MLflow, ensuring full
        data-to-model lineage.
  \item \textbf{Continuous monitoring}: live Twitter/X data is scraped daily, classified, and analysed
        for distribution shift against the training reference using PSI and KS-tests.
  \item \textbf{Automated retraining}: detected drift triggers a SLURM job on the TRUBA national HPC
        cluster via GitHub Actions self-hosted runner, without manual intervention.
  \item \textbf{Systematic comparison}: we evaluate five model families under identical conditions to
        provide a reproducible baseline progression from classical ML to transformer-based approaches.
\end{enumerate}

Our contributions are:
\begin{itemize}
  \item A fully automated, open-source fake news detection pipeline implementing MLOps best practices.
  \item A systematic benchmark comparing TF-IDF+LogReg, TF-IDF+SVM, LSTM, TextCNN, and BERTweet on
        FakeNewsNet and PHEME.
  \item A production-ready drift detection system based on PSI and KS statistics that integrates
        directly with HPC job scheduling.
  \item An empirical analysis of distribution shift between curated news datasets and live social
        media streams.
\end{itemize}


% ── 2. RELATED WORK ───────────────────────────────────────────────────────────
\section{Related Work}

\subsection{Fake News Detection}

Early approaches to automated fake news detection relied on handcrafted linguistic features combined
with classical classifiers~\cite{perez2017automatic}. TF-IDF representations paired with logistic
regression and SVM remain strong baselines, particularly on headline-level data~\cite{ahmed2018detecting}.

Deep learning approaches improved upon these baselines by capturing sequential and local dependencies.
LSTM-based models leverage temporal patterns in word sequences~\cite{hochreiter1997long}, while
Convolutional Neural Networks (CNNs) applied to text — notably the architecture of Kim~\cite{kim2014convolutional}
— capture local n-gram features through multiple filter widths. Both have been applied effectively to
fake news detection~\cite{wang2017liar}.

Pre-trained transformer models have recently dominated the field. BERT~\cite{devlin2019bert} and its
domain-specific variants, such as BERTweet~\cite{nguyen2020bertweet} (trained on 850M English tweets),
offer contextualised representations that are especially well-suited to social media text. Studies
applying BERTweet to rumour verification on PHEME~\cite{zubiaga2016analysing} consistently report
improvements over prior approaches.

\subsection{MLOps for NLP}

MLOps — the discipline of bringing DevOps principles to machine learning — has seen growing adoption
in industry but remains underexplored in the NLP research literature~\cite{sculley2015hidden}.
Key components include experiment tracking (MLflow~\cite{zaharia2018accelerating}), data and model
versioning (DVC~\cite{kupriianova2021dvc}), and automated CI/CD pipelines. \citet{klaise2021alibi}
demonstrate the importance of model monitoring for detecting input drift in production NLP systems.
To our knowledge, FakeShield is the first open-source fake news detection system that integrates
statistical drift detection with automated HPC-triggered retraining.


% ── 3. DATASETS ───────────────────────────────────────────────────────────────
\section{Datasets}

\subsection{FakeNewsNet}

FakeNewsNet~\cite{shu2020fakenewsnet} aggregates two political and celebrity news datasets collected
from PolitiFact and GossipCop. Each record contains a news headline and a binary label (real/fake).
We use the headline text exclusively, as full article URLs suffer from extensive link rot.
After deduplication and label filtering, our working corpus contains \textbf{22,775 articles}
(GossipCop: 18,290; PolitiFact: 4,485), split 80/20 into training and test sets with stratification.
The dataset is class-imbalanced, with real articles comprising approximately 74\% of samples.

\subsection{PHEME}

The PHEME dataset~\cite{zubiaga2016analysing} is a collection of rumourous Twitter threads organised
around nine breaking news events: Charlie Hebdo, Ferguson, Germanwings crash, Ottawa shooting,
Sydney siege, and others. Source tweets are annotated at the thread level as \textit{rumour} or
\textit{non-rumour}, with rumours further labelled as \textit{true}, \textit{false}, or
\textit{unverified}. Following standard practice~\cite{kochkina2018all}, we retain only verified
annotations, mapping \textit{false rumours} to label 0 (fake) and \textit{non-rumours} and
\textit{true rumours} to label 1 (real), discarding unverified instances.
The resulting dataset contains \textbf{2,006 labelled source tweets}, split 80/20 for fine-tuning
and evaluation of BERTweet.

\subsection{Live Twitter/X Stream}

For real-time monitoring, we continuously scrape US political discourse from Twitter/X via the
Twitter API v2. Queries target terms related to US domestic politics and geopolitical events
(e.g., \textit{Iran, sanctions, Pentagon, Congress}). As of the writing of this paper, we have
accumulated \textbf{194 unique tweets} across multiple daily scrape cycles, all English-language
and deduplicated by text. These tweets serve exclusively as inference targets for drift monitoring;
they are not used for training.


% ── 4. SYSTEM ARCHITECTURE ────────────────────────────────────────────────────
\section{System Architecture}

FakeShield is organised as a directed acyclic pipeline managed by DVC, with all components
integrated into a GitHub Actions CI/CD workflow. Figure~\ref{fig:pipeline} illustrates the
full system.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  node distance=0.6cm and 1.2cm,
  box/.style={rectangle, rounded corners, draw=black, fill=blue!10,
              text centered, minimum height=0.7cm, minimum width=2.2cm, font=\small},
  arrow/.style={->, >=stealth, thick},
  dasharrow/.style={->, >=stealth, thick, dashed}
]

\node[box] (ingest)   {Data Ingestion};
\node[box, right=of ingest] (prep) {Preprocessing};
\node[box, right=of prep]   (train) {Model Training};
\node[box, right=of train]  (eval)  {Evaluation};

\node[box, below=1.0cm of ingest] (scrape) {Live Scraper\\(Twitter/X)};
\node[box, right=of scrape]       (monitor){Drift Monitor\\(PSI + KS)};
\node[box, right=of monitor]      (ci)     {GitHub Actions\\CI/CD};
\node[box, right=of ci]           (truba)  {TRUBA HPC\\(SLURM)};

\draw[arrow] (ingest)  -- (prep);
\draw[arrow] (prep)    -- (train);
\draw[arrow] (train)   -- (eval);
\draw[arrow] (scrape)  -- (monitor);
\draw[dasharrow] (eval) -- node[right,font=\tiny]{reference} (monitor);
\draw[arrow] (monitor) -- node[above,font=\tiny]{drift?} (ci);
\draw[arrow] (ci)      -- node[above,font=\tiny]{sbatch} (truba);
\draw[dasharrow] (truba) to[bend right=40] node[below,font=\tiny]{push metrics} (eval);

\end{tikzpicture}
\caption{FakeShield pipeline architecture. Solid arrows indicate data flow; dashed arrows indicate
         reference distribution propagation and metric feedback.}
\label{fig:pipeline}
\end{figure}

\subsection{Data and Model Versioning}

All data transformations and model artefacts are tracked using DVC with a Git-backed metadata store.
The pipeline defines six stages: \texttt{ingest}, \texttt{preprocess}, \texttt{train},
\texttt{evaluate}, \texttt{baselines}, and \texttt{train\_bertweet}. Model binaries exceeding
GitHub's size threshold (50\,MB) are stored via Git LFS. Hyperparameters for all models are
centralised in \texttt{params.yaml}, ensuring full reproducibility.

\subsection{Experiment Tracking}

Every training run is logged to MLflow, capturing hyperparameters, per-epoch metrics, and final
evaluation scores. This enables systematic comparison across model families and facilitates
identification of the best checkpoint for deployment.

\subsection{Real-Time Inference Serving}

A FastAPI application (\texttt{src/app.py}) serves the production TextCNN model for both
interactive headline analysis and batch inference over the live tweet stream. The application
resolves all paths relative to the project root, ensuring portability across deployment environments.

\subsection{Drift Monitoring}

The monitoring module (\texttt{src/monitor.py}) runs daily within GitHub Actions and implements
two complementary drift detectors:

\textbf{Kolmogorov-Smirnov Test}: A two-sample KS test between the model's predicted probability
distribution on the live stream and the held-out training distribution. A $p$-value below 0.05
indicates statistically significant distribution shift.

\textbf{Population Stability Index (PSI)}: A symmetric divergence metric widely used in
production ML systems~\cite{yurdakul2018statistical}. PSI is computed over 10 equal-frequency bins
of the predicted probability distribution:
\[
  \mathrm{PSI} = \sum_{i=1}^{10} \left( A_i - E_i \right) \ln\!\frac{A_i}{E_i}
\]
where $A_i$ and $E_i$ are the actual and expected proportions in bin $i$. We adopt the
standard thresholds: $\mathrm{PSI} < 0.10$ (stable), $0.10$--$0.25$ (moderate drift),
$\geq 0.25$ (retrain).

Additionally, the monitor reports the US political relevancy rate (fraction of tweets matching
a curated keyword list) and average model confidence, which serve as early-warning indicators
of topic drift before statistical thresholds are reached.

\subsection{Automated HPC Retraining}

When drift is detected (monitor exits with code 1), the \texttt{scrape\_and\_monitor.yml}
workflow triggers \texttt{retrain\_truba.yml} via the GitHub Actions API. This workflow runs on
a self-hosted runner installed on the TRUBA login node and submits a SLURM batch job to the
\texttt{barbun-cuda} GPU partition (1 node, 1 GPU, 32\,GB RAM, 8 CPUs, 6-hour wall time).
The SLURM job downloads fresh PHEME data, preprocesses it, fine-tunes BERTweet, and pushes
the resulting metrics back to the GitHub repository, closing the retraining loop automatically.


% ── 5. MODELS ─────────────────────────────────────────────────────────────────
\section{Models}

\subsection{TF-IDF Baselines}

We construct bag-of-words representations using TF-IDF weighting with a vocabulary of 30,000
unigrams and bigrams. Two classifiers are evaluated: Logistic Regression (C=1.0) and a linear
SVM with Platt scaling (CalibratedClassifierCV) for probability calibration (C=1.0).

\subsection{LSTM}

A single-layer bidirectional LSTM with 128 hidden units processes a sequence of word embeddings
(128-dimensional, trained from scratch). A dense layer with sigmoid activation produces the
binary classification output. Training uses early stopping with patience 3 on validation F1.

\subsection{TextCNN}

Our primary deployment model follows the multi-filter CNN architecture of~\citet{kim2014convolutional}.
Three parallel convolutional layers with filter widths $\{3, 4, 5\}$ and 128 filters each operate
over 128-dimensional word embeddings. Max-over-time pooling aggregates each filter bank; the
concatenated representation (384-dimensional) is passed through dropout (0.5) and a dense layer
(128 units, ReLU) before the sigmoid output. The model is trained for up to 15 epochs with
early stopping (patience 3) using the Adam optimiser (lr=$10^{-3}$) on binary cross-entropy.

\subsection{BERTweet}

BERTweet~\cite{nguyen2020bertweet} is a RoBERTa-base model pre-trained on 850M English tweets,
making it the natural choice for the PHEME tweet domain. We fine-tune the \texttt{vinai/bertweet-base}
checkpoint for binary classification using the HuggingFace Transformers library. Training uses
AdamW ($\text{lr}=2\times10^{-5}$, weight decay 0.01) with a linear warmup over 10\% of steps,
batch size 32, for 5 epochs. The best checkpoint (by macro-F1 on validation) is saved to
\texttt{models/bertweet\_finetuned/}.


% ── 6. EXPERIMENTS ────────────────────────────────────────────────────────────
\section{Experiments}

\subsection{Setup}

All FakeNewsNet experiments use an identical 80/20 stratified train/test split with random seed 42.
Text is lowercased and stripped of punctuation and URLs. Evaluation reports accuracy, macro-averaged
precision, recall, and F1, per-class F1 for fake and real, and AUC-ROC. BERTweet is evaluated
separately on the PHEME dataset under the same split configuration, with tweet-specific normalisation
(URLs replaced with \texttt{HTTPURL}, mentions with \texttt{@USER}).

FakeNewsNet experiments run on a local CPU environment. BERTweet fine-tuning runs on TRUBA
(\texttt{barbun-cuda} partition, NVIDIA GPU, 32\,GB RAM).

\subsection{Results on FakeNewsNet}

Table~\ref{tab:results} reports the performance of all models on the FakeNewsNet test set.

\begin{table}[h]
\centering
\caption{Model performance on FakeNewsNet (headlines, 20\% test split). Best scores in \textbf{bold}.}
\label{tab:results}
\begin{tabular}{lcccccc}
\toprule
\textbf{Model} & \textbf{Acc.} & \textbf{Prec.} & \textbf{Rec.} & \textbf{F1} & \textbf{F1\textsubscript{fake}} & \textbf{AUC} \\
\midrule
TF-IDF + LogReg & 0.837 & 0.780 & 0.803 & 0.790 & 0.691 & \textbf{0.882} \\
TF-IDF + SVM    & \textbf{0.853} & \textbf{0.831} & 0.752 & 0.779 & 0.651 & 0.876 \\
LSTM            & 0.841 & 0.791 & 0.765 & 0.776 & 0.657 & 0.869 \\
TextCNN         & 0.832 & 0.774 & \textbf{0.801} & \textbf{0.785} & \textbf{0.686} & 0.876 \\
\bottomrule
\end{tabular}
\end{table}

TF-IDF+SVM achieves the highest accuracy (0.853) by learning a high-precision decision boundary,
but at the cost of recall on the fake class (F1\textsubscript{fake}=0.651). TextCNN, despite lower
overall accuracy, achieves the best fake-class F1 (0.686) and the best recall (0.801), suggesting
its multi-scale convolutional filters capture more diverse deceptive patterns. The strong
performance of TF-IDF+LogReg on AUC-ROC (0.882) confirms that bag-of-words features carry
substantial discriminative signal for headline-level misinformation.

Notably, all deep learning models (LSTM, TextCNN) slightly underperform TF-IDF baselines on overall
accuracy. We attribute this to the brevity of FakeNewsNet headlines (median 10 tokens), which limits
the advantage of sequential or local dependency modelling. This finding motivates the use of
PHEME tweet data for BERTweet fine-tuning, where longer and noisier text favours contextualised representations.

\subsection{Results on PHEME (BERTweet)}

\begin{table}[h]
\centering
\caption{BERTweet fine-tuned on PHEME tweet dataset.}
\label{tab:bertweet}
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \textbf{Acc.} & \textbf{F1\textsubscript{macro}} & \textbf{F1\textsubscript{fake}} & \textbf{AUC} \\
\midrule
BERTweet (fine-tuned) & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textit{Note: BERTweet results will be populated after completion of TRUBA training run.
Results are expected to exceed 0.85 macro-F1 based on prior work on PHEME~\cite{kochkina2018all}.}

\subsection{Discussion}

Several observations merit discussion.

\textbf{Class imbalance.} All models exhibit lower F1 on the fake class (F1\textsubscript{fake})
than the real class, reflecting the 74\%/26\% real/fake imbalance in FakeNewsNet. Future work
should explore class-weighted loss or oversampling strategies.

\textbf{Dataset limitations.} FakeNewsNet provides only headlines, not full article text.
Transformer models such as BERT are unlikely to demonstrate their full advantage on 10-token inputs.
The PHEME fine-tuning experiment addresses this by evaluating on full social media posts with
richer contextual signals.

\textbf{Domain shift.} The monitoring module reports an average confidence of 0.50 when classifying
live Twitter/X stream data with the FakeNewsNet-trained TextCNN. This near-random confidence
directly demonstrates domain mismatch: a model trained on news headlines cannot reliably classify
short political tweets. BERTweet fine-tuned on PHEME is designed to resolve this gap.


% ── 7. DRIFT MONITORING ───────────────────────────────────────────────────────
\section{Live Monitoring and Retraining Loop}

\subsection{Monitoring in Practice}

FakeShield has been running its daily monitoring pipeline for six days as of this writing,
accumulating 194 unique English-language political tweets. The current drift report shows
relevancy rate 1.0 (all scraped tweets match the US political keyword filter), confirming
the scraper is on-target. The KS and PSI values are currently null, pending population of the
reference distribution from the PHEME training split; this is resolved as part of the BERTweet
pipeline completion.

\subsection{Retraining Trigger Semantics}

The monitoring script exits with code 1 under three conditions: (1) PSI $\geq 0.25$,
(2) KS-test $p < 0.05$, or (3) average model confidence below 0.30 for more than three
consecutive days. Condition (3) serves as a safety net for the case where drift is gradual
and statistical tests lack power.

\subsection{HPC Integration}

The TRUBA self-hosted runner maintains a persistent outbound polling connection to GitHub's
Actions API from the login node, requiring no VPN tunnel from GitHub's cloud infrastructure.
The runner submits jobs to the SLURM scheduler on the \texttt{barbun-cuda} GPU partition and
pushes resulting metrics back to the repository upon job completion, making the retraining
outcome visible in the repository's metrics history without manual SSH access.


% ── 8. CONCLUSION ─────────────────────────────────────────────────────────────
\section{Conclusion}

We presented FakeShield, a production-ready MLOps pipeline for fake news detection that bridges
the gap between static benchmark evaluation and real-world deployment. The system provides a
reproducible multi-model comparison spanning classical ML, deep learning, and transformer
architectures, integrated with continuous drift monitoring and automated HPC retraining.
Our experiments on FakeNewsNet demonstrate that TF-IDF baselines remain competitive on
headline-level data, while the BERTweet fine-tuning on PHEME extends the system to the tweet
domain where transformer representations are most advantageous. FakeShield is publicly available
at \url{https://github.com/zegdane1998/fake-news-mlops}.

Future work includes: (1) evaluating additional transformer models (RoBERTa, DeBERTa-v3) under
the same pipeline; (2) extending monitoring to multilingual content; (3) implementing event-level
evaluation on PHEME to understand model generalisation across different breaking news scenarios;
and (4) applying active learning to efficiently label uncertain live tweets for incremental retraining.


% ── REFERENCES ────────────────────────────────────────────────────────────────
\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
